---
title: "Customizing the Format and the Destination of a Log Record"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customizing the Format and the Destination of a Log Record}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r pkgchecks, echo = FALSE}
## check if other logger packages are available and exit if not
for (pkg in c("devtools", "parallel")) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    warning(paste(pkg, "package not available, so cannot build this vignette"))
    knitr::knit_exit()
  }
}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(logger)
log_appender(appender_stdout)
```

## Anatomy of a request

```{r loggerStructureImage, echo=FALSE, out.extra='style="width: 100%;" title="The structure of a logger and the flow of a log record request" alt="The structure of a logger and the flow of a log record request"'}
knitr::include_graphics("logger_structure.svg")
```

To make a successful log record, `logger` requires the below components:

- a **log request**, eg

    ```r
    log_error('Oops')
    ```

    - including the log level (importance) of the record, which will be later used to decide if the log record is to be delivered or not: `ERROR` in this case
    - R objects to be logged: a simple string in this case, although it could be a character vector or any R object(s) that can be converted into a character vector by the `formatter` function

- the **environment** and meta-information of the log request, eg actual timestamp, hostname of the computer, the name of the user running the R script, the pid of the R process, calling function and the actual call etc.


    ```{r}
f <- function() get_logger_meta_variables(log_level = INFO)
f()
    ```

- a **logger definition** to process the log request, including

      - log level `threshold`, eg `INFO`, which defines the minimum log level required for actual logging -- all log requests with lower log level will be thrown away

      ```{r}
      log_threshold()
      ERROR <= INFO
      log_error("Oops")
      ```

      - `formatter` function, which takes R objects and converts those into actual log message(s) to be then passed to the `layout` function for the log record rendering -- such as `paste`, `sprintf`, `glue` or eg the below custom example:


      ```{r}
      formatter <- function(...) paste(..., collapse = " ", sep = " ")
      formatter(1:3, c("foo", "bar"))
      ```

     - `layout` function, which takes log message(s) and further information on the log request (such as timestamp, hostname, username, calling function etc) to render the actual log records eg human-readable text, JSON etc

      ```r
      library(jsonlite)
      layout <- function(level, msg) toJSON(level = level, timestamp = time, hostname = node, message = msg)
      layout(INFO, 'Happy Thursday!')
      #> {'level': 'INFO', 'timestamp': '1970-01-01 00:00:00', 'hostname': 'foobar', 'message': 'Happy Thursday!'}
      ```

      - `appender` function, which takes fully-rendered log record(s) and delivers to somewhere, eg `stdout`, a file or a streaming service, eg

      ```{r}
      appender <- function(line) cat(line, "\n")
      appender("INFO [now] I am a log message")
      ```

Putting all these together (by explicitly setting the default config in the `global` namespace):

```{r}
log_threshold(INFO)
log_formatter(formatter_glue)
log_layout(layout_simple)
log_appender(appender_stdout)
log_debug("I am a low level log message that will not be printed with a high log level threshold")
log_warn("I am a higher level log message that is very likely to be printed")
```

Note, that all `logger` definitions and requests are tied to a logging namespace, and one log request might trigger multiple `logger` definitions as well (stacking). Find more information on these in the `vignette("customize_logger")`.

## What gets logged?

`logger` mostly relies on and uses the default `log4j` log levels and supports suppressing log messages with a lower log level compared to the currently set threshold in the logging namespace:

```{r}
log_info("Hi, there!")
log_debug("How are you doing today?")
log_threshold()
log_threshold(TRACE)
log_debug("How are you doing today?")
```

So the `?log_threshold` function can both get and set the log level threshold for all future log requests.

For the full list of all supported log levels and so thus the possible log level thresholds, see `?log_levels`.

If you want to define the log level in a programmatic way, check out `?log_level`, eg

```{r}
log_level(INFO, "Hi, there!")
```

To temporarily update the log level threshold, you may also find the `?with_log_threshold` function useful:

```{r}
log_threshold(INFO)
log_debug("pst, can you hear me?")
log_info("no")

with_log_threshold(log_debug("pst, can you hear me?"), threshold = TRACE)
log_info("yes")

with_log_threshold(
  {
    log_debug("pst, can you hear me?")
    log_info("yes")
  },
  threshold = TRACE
)
```

You can also define your own log level(s) if needed, for example introducing an extra level between `DEBUG` and `INFO`:

```{r}
FYI <- structure(450L, level = "FYI", class = c("loglevel", "integer"))
log_threshold(FYI)
log_debug("ping")
log_level(FYI, "ping")
log_info("pong")
```

## Log namespaces

By default, all log messages will be processed by the global `logger` definition, but you may also use custom namespaces (eg to deliver specific log records to a special destination or to apply a custom log level threshold) and even multiple loggers as well within the very same namespace (eg to deliver all `INFO` and above log levels in the console and everything below that to a trace log file).

If you specify an unknown `namespace` in a log request, it will fall back to the global settings:

```{r}
log_threshold(INFO)
log_trace("Hi, there!", namespace = "kitchensink")
log_info("Hi, there!", namespace = "kitchensink")
```

But once you start customizing that namespace, it gets forked from the global settings and live on its own without modifying the original namespace:


```{r}
log_threshold(TRACE, namespace = "kitchensink")
log_trace("Hi, there!", namespace = "kitchensink")
log_info("Hi, there!", namespace = "kitchensink")
log_trace("Hi, there!")
```

## Log message formatter functions

In the above example, we logged strings without any dynamic parameter, so the task of the logger was quite easy. But in most cases you want to log a parameterized string and the formatter function's task to transform that to a regular character vector.

By default, `logger` uses `glue` in the background:

```{r}
log_formatter(formatter_glue)
log_info("There are {nrow(mtcars)} cars in the mtcars dataset")
log_info("2 + 2 = {2+2}")
```

If you don't like this syntax, or want to save a dependency, you can use other formatter functions as well, such as `formatter_sprintf()` (the default in eg the `logging` and `futile.logger` packages, which you can learn more about in `vignette("migration.html")`)  or `formatter_paste()`, or you can write your own, as described in `vignette("write_custom_extensions.html")`.

## Log message layouts

By default, `?log_level` and its derivative functions (eg `?log_info`) will simply record the log-level, the current timestamp and the message after being processed by `glue`:

```{r}
log_info(42)
log_info("The answer is {42}")
log_info("The answers are {1:5}")
```

In the above example, first, `42` was converted to a string by the `?formatter_glue` message formatter, then the message was passed to the `?layout_simple` layout function to generate the actual log record.

An example of another layout function writing the same log messages in JSON:

```{r}
log_layout(layout_json())
log_info(42)
log_info("The answer is {42}")
log_info("The answers are {1:5}")
```

If you need colorized logs highlighting the important log messages, check out `?layout_glue_colors`, and for other formatter and layout functions, see the manual of the above mentioned functions that have references to all the other functions and generator functions bundled with the package.

## Custom log record layout

To define a custom format on how the log messages should be rendered, you may write your own `formatter` and `layout` function(s) or rely on the function generator functions bundled with the `logger` package, such as `?layout_glue_generator`.

This function returns a `layout` function that you can define by `glue`-ing together variables describing the log request via `?get_logger_meta_variables`, so having easy access to (package) namespace, calling function's name, hostname, user running the R process etc.

A quick example:

* define custom logger:

    ```{r}
logger <- layout_glue_generator(format = "{node}/{pid}/{namespace}/{fn} {time} {level}: {msg}")
log_layout(logger)
    ```

* check what's being logged when called from the global environment:

    ```{r}
log_info("foo")
    ```

* check what's being logged when called from a custom function:

    ```{r}
f <- function() log_info("foo")
f()
    ```

* check what's being logged when called from a package:

    ```{r}
devtools::load_all(system.file("demo-packages/logger-tester-package", package = "logger"))
logger_tester_function(INFO, "hi from tester package")
    ```

* suppress messages in a namespace:

    ```{r}
log_threshold(namespace = "logger.tester")
log_threshold(WARN, namespace = "logger.tester")
logger_tester_function(INFO, "hi from tester package")
logger_tester_function(WARN, "hi from tester package")
log_info("I am still working in the global namespace")
    ```

Another example of making use of the generator function is to update the layout to include the Process ID that might be very useful eg when forking, see for example the below code chunk still using the above defined log layout:

```r
f <- function(x) {
    log_info('received {length(x)} values')
    log_success('with the mean of {mean(x)}')
    mean(x)
}
library(parallel)
mclapply(split(runif(100), 1:10), f, mc.cores = 5)
#> nevermind/26448/R_GlobalEnv/FUN 2018-12-02 21:54:11 INFO: received 10 values
#> nevermind/26448/R_GlobalEnv/FUN 2018-12-02 21:54:11 SUCCESS: with the mean of 0.403173440974206
#> nevermind/26449/R_GlobalEnv/FUN 2018-12-02 21:54:11 INFO: received 10 values
#> nevermind/26448/R_GlobalEnv/FUN 2018-12-02 21:54:11 INFO: received 10 values
#> nevermind/26449/R_GlobalEnv/FUN 2018-12-02 21:54:11 SUCCESS: with the mean of 0.538581100990996
#> nevermind/26448/R_GlobalEnv/FUN 2018-12-02 21:54:11 SUCCESS: with the mean of 0.485734378430061
#> nevermind/26450/R_GlobalEnv/FUN 2018-12-02 21:54:11 INFO: received 10 values
#> nevermind/26449/R_GlobalEnv/FUN 2018-12-02 21:54:11 INFO: received 10 values
#> nevermind/26450/R_GlobalEnv/FUN 2018-12-02 21:54:11 SUCCESS: with the mean of 0.580483326432295
#> nevermind/26452/R_GlobalEnv/FUN 2018-12-02 21:54:11 INFO: received 10 values
#> nevermind/26449/R_GlobalEnv/FUN 2018-12-02 21:54:11 SUCCESS: with the mean of 0.461282140854746
#> nevermind/26450/R_GlobalEnv/FUN 2018-12-02 21:54:11 INFO: received 10 values
#> nevermind/26451/R_GlobalEnv/FUN 2018-12-02 21:54:11 INFO: received 10 values
#> nevermind/26450/R_GlobalEnv/FUN 2018-12-02 21:54:11 SUCCESS: with the mean of 0.465152264293283
#> nevermind/26452/R_GlobalEnv/FUN 2018-12-02 21:54:11 SUCCESS: with the mean of 0.618332817289047
#> nevermind/26451/R_GlobalEnv/FUN 2018-12-02 21:54:11 SUCCESS: with the mean of 0.493527933699079
#> nevermind/26452/R_GlobalEnv/FUN 2018-12-02 21:54:11 INFO: received 10 values
#> nevermind/26452/R_GlobalEnv/FUN 2018-12-02 21:54:11 SUCCESS: with the mean of 0.606248055002652
#> nevermind/26451/R_GlobalEnv/FUN 2018-12-02 21:54:11 INFO: received 10 values
#> nevermind/26451/R_GlobalEnv/FUN 2018-12-02 21:54:11 SUCCESS: with the mean of 0.537314630229957
```

*Note that the `layout_glue_generator` functions also adds a special attribute to the resulting formatting function so that when printing the layout function to the console, the user can easily interpret what's being used instead of just showing the actual functions's body:*

```{r}
log_layout()
```

For more details on this, see `vignette("write_custom_extensions.html")`.

```{r}
## reset layout
log_layout(layout_simple)
```

## Delivering log records to their destination

By default, `logger` will write to the `stderr` via the `?appender_console` function:

```{r}
log_appender()
```

To write to a logfile instead, use the `?appender_file` generator function, that returns a function that can be used in any namespace:

```{r}
t <- tempfile()
log_appender(appender_file(t))
log_info("where is this message going?")
log_appender()
readLines(t)
unlink(t)
```

There's a similar generator function that returns an appender function delivering log messages to Slack channels:

```r
## load Slack configuration, API token etc from a (hopefully encrypted) yaml file or similar
slack_config <- config::config(...)
## redirect log messages to Slack
log_appender(appender_slack(
    channel   = '#gergely-test',
    username  = 'logger',
    api_token = slack_config$token
), namespace = 'slack')
log_info('Excited about sending logs to Slack!')
#> INFO [2018-11-28 00:21:13] Excited about sending logs to Slack!
log_info('Hi there from logger@R!', namespace = 'slack')
```

You may find `?appender_tee` also useful, that writes the log messages to both `stdout` and a file.

```{r}
## reset appender
log_appender(appender_stdout)
```

And the are many other appender functions bundled with `logger` as well, eg some writing to Syslog, Telegram, Pushbullet, a database table or an Amazon Kinesis stream -- even doing that asynchronously via `appender_async` -- see `vignette("performance")` for more details.

## Stacking loggers

Note that the `?appender_tee` functionality can be implemented by stacking loggers as well, eg setting two loggers for the global namespace: `?appender_console` and `?appender_file`. The advantage of this approach is that you can set different log level thresholds for each logger, for example:

```{r}
log_threshold()

## create a new logger with index 2
log_threshold(TRACE, index = 2)

## note that the original logger still have the same log level threshold
log_threshold()
log_threshold(index = 2)

## update the appender of the new logger
t <- tempfile()
log_appender(appender_file(t), index = 2)

## test both loggers
log_info("info msg")
log_debug("info msg")

readLines(t)
unlink(t)
```

```{r cleanup, include = FALSE}
logger:::namespaces_reset()
```
