---
title: "Migration Guide"
vignette: >
  %\VignetteIndexEntry{Migration Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
div.left, div.right {
    width: 49%;
    display: inline-block;
    vertical-align: top;
}
div.left p package, div.right p package {
    font-family: monospace;
    color: brown;
}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette I suppose that you are already familiar with at least one of the [similar logging R packages](https://daroczig.github.io/logger/index.html#why-another-logging-r-package) and you are looking for suggestions on how to switch to `logger`.

## futile.logger

## logging

The `logging` package behaves very similarly to the Python logging module and so thus being pretty Pythonic, while `logger` tries to accommodate native R users' expectations -- so there are some minor nuances between the usage of the two packages.

In `logging`, you have to initialize a logger first via `addHandler` or simply by calling `basicConfig`, which is not required in `logger` as it already comes with a default logger.

<div class="left">
<package>logging</package>
```{r}
library(logging)
basicConfig()
```
</div>
<div class="right">
<package>logger</package>
```{r}
library(logger)
```
</div>

Logging works similarly in the two packages -- with a bit different function names:

* although `logging` uses mostly camelCase function names (eg `basicConfig`), but the logging functions are all lowercase without any separator, such as `loginfo` or `logwarn`
* `logger` uses snake_case for the function names, such as `log_info` and `log_warn`

<div class="left">
<package>logging</package>
```{r}
1+2
```
</div>
<div class="right">
<package>logger</package>
```{r}
1+2
```
</div>

+--------------------------+---------------------------+
| logging                  | logger                    |
+==========================+===========================+
|     loginfo('hi there')  |     log_info('hi there')  |
|     logwarn('watch out') |     log_warn('watch out') |
+--------------------------+---------------------------+

The default output is also a bit different:

+--------------------------+---------------------------+
| logging                  | logger                    |
+==========================+===========================+
|     loginfo('hi there')  |     log_info('hi there')  |
|     logwarn('watch out') |     log_warn('watch out') |
+--------------------------+---------------------------+



